\documentclass[svgnames]{article}     % use "amsart" instead of "article" for AMSLaTeX format
%\geometry{landscape}                 % Activate for rotated page geometry

%\usepackage[parfill]{parskip}        % Activate to begin paragraphs with an empty line rather than an indent

\usepackage{graphicx}                 % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode

%maths                                % TeX will automatically convert eps --> pdf in pdflatex
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{esint}
\usepackage{geometry}

% Inverting Color of PDF
\usepackage{xcolor}
\pagecolor[rgb]{0.19,0.19,0.19}
\color[rgb]{0.77,0.77,0.77}

%noindent
\setlength\parindent{0pt}

%pgfplots
\usepackage{pgfplots}

%images
%\graphicspath{{ }}                   % Activate to set a image directory

%tikz
\usepackage{pgfplots}
\pgfplotsset{compat=1.15}
\usepackage{comment}
\usetikzlibrary{arrows}
\usepackage[most]{tcolorbox}

\usepackage{listings}
\usepackage{color}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
  language=Java,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}

\lstset{language=Python}

\tcbset {
  base/.style={
    arc=0mm,
    bottomtitle=0.5mm,
    boxrule=0mm,
    colbacktitle=black!10!white,
    coltitle=black,
    fonttitle=\bfseries,
    left=2.5mm,
    leftrule=1mm,
    right=3.5mm,
    title={#1},
    toptitle=0.75mm,
  }
}

\definecolor{brandblue}{rgb}{0.34, 0.7, 1}
\newtcolorbox{mainbox}[1]{
  colframe=brandblue,
  base={#1}
}

\newtcolorbox{subbox}[1]{
  colframe=black!30!white,
  base={#1}
}

%Figures
\usepackage{float}
\usepackage{caption}
\usepackage{lipsum}


\title{Multiple Systems -- Lesson 02}
\author{Deval Deliwala}
%\date{}                              % Activate to display a given date or no date

\begin{document}
\maketitle
%\section{}
%\subsection{}
\tableofcontents                     % Activate to display a table of contents
\newpage

We can always choose to view multiple systems \textit{together} as if they form
a single, compound system -- to which the discussion in the previous section
applies. 

\section{Classical Information}

Recognizing how mathematics works in the familiar setting of classical
information is helpful in understanding why quantum information is described
the way it is. 

\subsection{Classical States via the Cartesian Product}

We will start with classical states of multiple systems. We will begin by
discussing just two systems and then generalize further. \\

To be precise, let us suppose that $X$ is a system whose classical state set is
$\Sigma$ and $Y$ is a second system having classical state set $\Gamma$.
Because these are \textit{classical state sets}, our assumption is that
$\Sigma$ and $\Gamma$ are both finite and nonempty. It could be that $\Sigma
= \Gamma$, but this is not necessarily so. \\

Now imagine the two systems  $X$ and $Y$ are placed side-by-side, with $X$ on the left and $Y$ on the right. We can views these two systems as if they form a single system $(X, Y)$ or $XY$. \\

The set of classical states of $XY$ is the \textit{Cartesian Product} of $\Sigma$ and $\Gamma$, which is the set defined as 

\begin{mainbox}{Cartesian Product}
\[
  \Sigma \times \Gamma = { (a, b) : a \in \Sigma \text{ and } b \in \Gamma}
\] 
\end{mainbox}

The Cartesian Product is the mathematical notion of viewing an element of one set and an element of a second set together, as if they form a single set. \\ 

For more than two systems, the situation generalizes in a natural way. If we suppose that $X_1, \cdots, X_n$ are systems having classical state sets $\Sigma_1, \cdots, \Sigma_n$, respectively for any positive integer $n$, the classical state set of the $n$-tuple $X_1X_2\cdots X_n$, viewed as a single joint system, is the Cartesian Product

\[
  \Sigma_1 \times \cdots \times \Sigma_n = {(a_1, \cdots, a_n) : a_1 \in \Sigma_1 \cdots a_n \in \Sigma_n}
\] \vspace{5px}

\subsection{Representing states as strings} 

It is convenient to write a classical state $(a_1, \cdots, a_n)$ as a string `$a_1 \cdots a_n$`. The notion of a string, which is fundamentally important concept in CS, is formalized in mathematical terms through the cartesian product. \\

For example suppose that $X_1, \cdots , X_{10}$ are bits, so the classical state sets of these systems are all the same 

\[
  \Sigma_1 = \Sigma_2 = \cdots = \Sigma_{10} = {0, 1}
\] \vspace{5px}

The set ${0, 1}$ is referred to as the \textit{binary alphabet}. There are
$2^{10} = 1024$ classical states of the joint system $X_1X_2 \cdots X_{10}$
which are elements of the set

\[
  \Sigma_1 \times \Sigma_2 \times \cdots \times \Sigma_{10} = {0, 1}^10. 
\] \vspace{5px}

Written as strings, these classical states look like this: 

\begin{align*}
  00000&00000 \\
  00000&00001 \\
  00000&00010 \\
  00000&00011 \\
  00000&00100 \\
       &\vdots \\
  11111&11111
\end{align*}

For the classical state $0001010000$ we see that $X_4$ and $X_6$ are in the
state 1, while all other systems are in the state 0. 

\subsection{Probabilistic States}

\textit{Probabilistic} states associate a probability with each classical state
of a system. Thus, a probabilistic state of multiple systems -- viewed
collectively as if they form a single system -- associates a probability with
each element of the Cartesian product of the classical state sets of the
individual systems. \\

Suppose $X$ and $Y$ are both bits. So their corresponding classical state sets
are $\Sigma = {0, 1}$ and  $\Gamma = {0, 1}$. Here is a probabilistic state of
the pair  $XY$: 

 \begin{align*}
   &\text{Pr}((X,Y) = (0,0)) = \frac{1}{2} \\
   &\text{Pr} ((X,Y) = (0, 1) = 0 \\
   &\text{Pr} ((X,Y) = (1, 0) = 0 \\
   &\text{Pr} ((X,Y) = (1, 1) = \frac{1}{2}
\end{align*}

This probabilistic state is one in which both $X$ and $Y$ are random bits --
each is 0 with probability 1/2 and 1 with probability 1/2 -- but the classical
states of the two bits always agree. This is an example of
a \textit{correlation} between these systems.

\subsection{Ordering Cartesian Product State Sets}

To represent a probabilistic state of multiple systems as a Cartesian product,
one must decide on an ordering of the product's elements. The entries in each
$n-$ tuple are viewed as being ordered by significance that \textit{decreases
from left to right}. \\

The Cartesian Product ${1, 2, 3} \times {0, 1}$ is ordered like this: 

 \[
   (1, 0), (1, 1), (2, 0), (2, 1), (3, 0), (3, 1)
\] \vspace{5px}

The probabilistic state represented by the probability vector above 

\[
\begin{pmatrix}
  \frac{1}{2}\\ 0 \\ 0 \\ \frac{1}{2}
\end{pmatrix}
\] \vspace{5px}

Where from top to bottom we get the probability associated with 00, then 01,
then 10, then 11. 

\subsection{Independence of Two Systems}

Intuitively speaking, two systems are independent if learning the classical
state of either system has no effect on the probabilities associated with the
other. \\

Suppose $X$ and $Y$ are systems having classical state sets $\Sigma$ and
$\Gamma$ respectively. They are said to be \textit{independent} if it is the
case that 

\[
\text{Pr}((X, Y) = (a, b)) = \text{Pr}(X = a)\text{Pr}(Y=b)
\] \vspace{5px}
for every $a \in \Sigma$ and $b \in \Gamma$. \\

To express this in terms of probability vectors, assume the probabilistic state
of $(X, Y)$ is described as a probability vector written in Dirac notation as 

\[
  \sum_{(a,b)\in \Sigma \times \Gamma} p_{ab} |ab\rangle
\] \vspace{5px}

The above condition is equivalent to the existence of two probability vectors 

\[
  |\phi\rangle = \sum_{a\in\Sigma} q_a |a\rangle \quad \text{and } \quad
  |\psi\rangle = \sum_{b\in\Gamma} r_b|b\rangle
\] \vspace{5px}

representing the probabilities associated with the classical states of $X$ and
$Y$ such that 

\[
  p_{ab} = q_a r_b
\] \vspace{5px}

For example, the probabilistic state of a pair of bits $(X,Y)$ represented by
the vector 

\[
\frac{1}{6}|00\rangle + \frac{1}{12}|01\rangle + \frac{1}{2}|10\rangle
+ \frac{1}{4}|11\rangle
\] \vspace{5px}

is one in which $X$ and $Y$ are independent. Specifically the condition
required for independence is true for the probability vectors 

\[
|\phi\rangle = \frac{1}{4}|0\rangle + \frac{3}{4}|1\rangle \quad \text{and}
\quad |\psi\rangle = \frac{2}{3}|0\rangle + \frac{1}{3}|1\rangle
\] \vspace{5px}

For example, to match the 00 entry we need $\frac{1}{6} = \frac{1}{4} \times
\frac{2}{3}$. The other entries can be verified in a similar manner.\\

On the other hand, the probabilistic state 

\[
\frac{1}{2}|00\rangle + \frac{1}{2}|11\rangle
\] \vspace{5px}

does not represent independence between $X$ and $Y$. \\

Suppose that there did exist probability vectors $|\phi\rangle$ and $\psi$ for
which $p_ab = q_ar_b$ is satisfied for every choice $a$ and $b$. It would then
necessarily be that 

\[
q_0r_1 = \text{Pr} ((X,Y) = (0, 1)) = 0
\] \vspace{5px}

This implies that either $q_0 = 0$ or $r_1 = 0$. Which means that $q_0r_0 = 0$
or $q_1r_1 = 0$ which isn't the case. Hence, there do not exist vectors $\phi$
and $|\psi\rangle$ satisfying the property required for independence. \\

We can now define correlation precisely as a \textit{lack of independence}. 

\subsection{Tensor Products of Vectors}

The condition for independence just described can be expressed more succinctly
through the notion of a \textit{tensor product}. In concrete terms, given two
vectors 

\[
  |\phi\rangle = \sum_{a\in\Sigma} \alpha_a |a\rangle \quad \text{and} \quad  |\psi\rangle = \sum_{b\in\Gamma} \beta_b |b\rangle
\] \vspace{5px}
the tensor product $|\phi\rangle \otimes |\psi\rangle $ is a new vector over
the joint set $\Sigma \times \Gamma$ defined as 
\begin{subbox}{Tensor Product}
\[
  |\phi\rangle \otimes |\psi\rangle = \sum_{(a,b) \in \Sigma \times \Gamma}
  \alpha_a \beta_b |ab\rangle 
\] 
\end{subbox}

Equivalently the vector $|\pi\rangle = |\phi\rangle \otimes |\psi\rangle $ is
defined by the equation 

\[
  \langle ab | \pi \rangle = \langle a | \phi \rangle \langle b | \psi \rangle
\] \vspace{5px}

being true for every $a\in \Sigma$ and $b\in\Gamma$

In this situation it is said that $|\pi\rangle$ is a \textit{product state} or
\textit{product vector}. We often just write $|\phi\rangle |\psi\rangle
$ instead of $|\phi\rangle \otimes |\psi\rangle $. The tensor product for two
column vectors is defined as follows: 

\[
\begin{pmatrix}
  \alpha_1 \\ \vdots \\ \alpha_m
\end{pmatrix} \otimes \begin{pmatrix}
  \beta_1 \\ \vdots \\ \beta_k 
\end{pmatrix} = \begin{pmatrix}
  \alpha_1\beta_1 \\ \vdots \\ \alpha_1\beta_k \\ \alpha_2 \beta_1 \\ \vdots \\
  \alpha_2 \beta_k \\ \vdots \\ \alpha_m \beta_1 \\ \vdots \\ \alpha_m \beta_k
\end{pmatrix} 
\] \vspace{5px}
For standard basis vectors: 

\[
  |a\rangle \otimes |b\rangle = |ab\rangle = |a, b\rangle
\] \vspace{5px}


The tensor product of two vectors has the important property that it is
\textit{bilinear}, which means that it is linear in each of the two arguments
separately, assuming that the other argument is fixed. 

\begin{mainbox}{Bilinearity}
  1. Linearity in the first argument: 

  \[
    (|\phi_1\rangle + |\phi_2\rangle) \otimes |\psi\rangle = |\phi_1\rangle
    \otimes |\psi\rangle + |\phi_2\rangle \otimes |\psi\rangle 
  \]\[
    ( \alpha | \phi \rangle) \otimes |\psi\rangle = \alpha (|\phi\rangle
    \otimes |\psi\rangle )
  \] 
  2. Linearity in the second argument: 

  \[
  |\phi\rangle \otimes (|\psi_1\rangle + |\psi_2\rangle ) = |\phi\rangle
\otimes |\psi_1\rangle  + |\phi\rangle \otimes |\psi_2\rangle \] \[
|\phi\rangle \otimes (\alpha |\psi\rangle) = \alpha(|\phi\rangle \otimes
|\psi\rangle )
\] 
\end{mainbox} 

Scalars ``float freely" within tensor products.

\subsection{Independence and Tensor Products for 3 or More Systems}

If $X_1, \cdots, X_n$ are systems with classical state sets $\Sigma_1, \cdots,
\Sigma_n$, then a probabilistic state of the combined system $X_1X_2\cdots
X_n$ is a \textit{product state}

 \[
|\psi\rangle = |\phi_1\rangle \otimes \cdots \otimes |\phi_n\rangle 
 \] \vspace{5px}
 
 
for probability vectors $|\phi_1\rangle \cdots |\phi_n\rangle $ describing
probabilistic states $X_1, \cdots, X_n$ \\
  
  
The definition of the tensor product generalizes as the following: 

\[
|\psi\rangle = |\phi_1\rangle \otimes \cdots \otimes |\phi_n\rangle 
\] \vspace{5px}

is defined by the equation 

\begin{subbox}{Tensor Product Definition}
\[
\langle a_1\cdots a_n | \psi \rangle = \langle a_1 | \phi_1\rangle \cdots
\langle a_n | \phi_n \rangle
\] \vspace{5px}
for every $a_1 \in \Sigma_1, \cdots a_n \in \Sigma_n$. 
\end{subbox}

Similar to the tensor product for two vectors, the tensor product for 3 or more
vectors is linear in each of the arguments individually. We say the tensor
product of 3 or more vectors is \textit{multilinear}. \\

For multiple systems is $X_1, ... X_n$ are \textit{independenet}, when they are
in a product state, they are actually \textit{mutually independent}. There is
also other notions of independence for 3 or more systems such as
\textit{pairwise independence}. \\

For any positive integer $n$ and any classical states $a_1, ... , a_n$ we have

\[
|a_1\rangle \otimes \cdot \otimes |a_n\rangle = |a_1\cdot a_n\rangle 
\] \vspace{5px}


\subsection{Measurements of Probabilistic States}

If two bits $(X, Y)$ are described by the following probability vector 

\[
\frac{1}{2}|00\rangle + \frac{1}{2}|11\rangle , 
\] \vspace{5px}

then the outcome measurement of 0 for $X$ and 0 for $Y$ is obtained with
probability $1/2$ and the outcome 11  is also obtained with probability $1/2$.
\\

Suppose however, that we choose not to measure \textit{every} system, just some
\textit{proper subset} of the systems. This will result in a measurement
outcome for each system that gets measured, and will also affect our knowledge
of the remaining systems. \\ 

Suppose that $X$ is a system having classical state set $\Sigma$ and  $Y$ is
a system having classical state set $\Gamma$, and the two systems together are
in some probabilistic state. We will consider what happens when we just measure
$X$ and do nothing with $Y$. The situation vice versa is handled symmetrically.
\\

We know that the probability to observe a particular classical state
$a\in\Sigma$ when just $X$ is measured must be consistent with the
probabilities we would obtain under the assumption that $Y$ was also measured.
That is, we must have

\[
  \text{Pr} (X = a) = \sum_{b\in\Gamma} \text{Pr} ((X, Y) = (a, b))
\] \vspace{5px}

Intuitively, if this formula is wrong, it would mean that the probabilities for
$X$ measurements are influenced simply by whether or not $Y$ is also measured,
irrespective of the outcome of $Y$. If $Y$ happened to be in a distant
location, say, another galaxy, this would allow for faster-than-light
signaling, which we obviously reject. \\

Given the assumption that only $X$ is measured, there still exists uncertainty
over the classical state of $Y$. We must update our description so this
uncertainty of  $Y$ is properly reflected. \\

The following \textit{conditional probability} formula reflects this
uncertainty. 

\[
\text{Pr} (Y = b \, | \, X = a) = \frac{\text{Pr} ((X,Y) = (a, b))}{\text{Pr} (X=a)}
\] \vspace{5px}

Here, the expression $\text{Pr} (Y=b \, | \, X = a )$ denotes the probability
that $Y = b$ \textit{conditioned} on $X = a$. \\

To express these formulas in terms of probability vector, consider
a probability vector $|\psi\rangle $ describing the joint state of $(X,Y)$. \\

\[
  |\psi\rangle = \sum_{(a,b) \in \Sigma \times \Gamma} p_{ab} |ab\rangle 
\] \vspace{5px}
Measuring $X$ alone yields each possible outcome with probabilities 

\[
  \text{Pr} (X=a) = \sum_{b\in\Gamma} p_{ab}
\] \vspace{5px}

Thus the vector representing the probabilistic state of $X$ alone is given by 

\[
  \sum_{a\in\Sigma} \left( \sum_{c\in\Gamma} p_{ac} \right) |a\rangle 
\] \vspace{5px}



The updated probabilistic state of $Y$ having obtained a particular outcome
$a\in\Sigma$ of the measurement of $X$, is represented by the probability
vector: 

\[
  |\pi_a\rangle = \frac{\sum_{b\in\Gamma} p_{ab}|b\rangle }{\sum_{c\in\Gamma}
  p_{ac}}
\] \vspace{5px}

If $X$ results in a classical state $a$, we update our description of the
probabilistic state of the joint system $(X,Y)$ to $|a\rangle \otimes
|\pi_a\rangle$. \\

The definition of $|\pi_a\rangle $ is the \textit{normalization} of the vector
$\sum_{b\in\Gamma} p_{ab}|b\rangle $ where we divide the sum of the entries in
this vector to obtain the probability vector. \\

\subsection{Example of Classical State Measurements}

For example, suppose the classical state of $X$ is $\Sigma = {0,1}$, the
classical state of  $Y$ is $\Gamma = {1, 2, 3}$, and the probabilistic state of
$(X, Y)$ is 

\[
|\psi\rangle  = \frac{1}{2}|0, 1\rangle + \frac{1}{12}|0,3\rangle
+ \frac{1}{12}|1,1\rangle + \frac{1}{6}|1, 2\rangle + \frac{1}{6}|1,3\rangle 
\] \vspace{5px}
Our goal is to determine probabilities of two possible outcomes -- 0 and 1, and
to calculate what the resulting probabilistic state of $Y$ is for the two
outcomes, assuming $X$ is measured. \\

Using the bilinearity of the tensor product, we may rewrite $|\psi\rangle $ as 

\[
|\psi\rangle = |0\rangle \otimes \left( \frac{1}{2}|1\rangle
  + \frac{1}{12}|3\rangle \right) + |1\rangle \otimes
  \left(\frac{1}{12}|1\rangle + \frac{1}{6}|2\rangle + \frac{1}{6}|3\rangle
    \right)
\] \vspace{5px}
We have isolated the distinct standard basis vectors for the system being
measured. This is always possible, regardless of what vector we start with. \\

Thus the probabilities of the two outcomes are given by 

\begin{align*}
&\text{Pr} (X=0) = \frac{1}{2} + \frac{1}{12} = \frac{7}{12} \\[5px]
&\text{Pr} (X=1) = \frac{1}{12} + \frac{1}{6} + \frac{1}{6} = \frac{5}{12}
\end{align*}

The probabilistic state of $Y$, conditioned on each possible outcome can also
be quickly inferred by normalizing the vectors in parenthesis so that these
vectors become probability vectors. That is, conditioned on $X$ being 0, the
probabilistic state of $Y$ becomes

\[
\frac{\frac{1}{2}|1\rangle + \frac{1}{12}|3\rangle }{\frac{7}{12}}
= \frac{6}{7}|1\rangle + \frac{1}{7}|3\rangle 
\] \vspace{5px}

and conditioned on the measurement of $X$ being 1, the probabilistic state of
$Y$ becomes


\[
\frac{\frac{1}{12}|1\rangle + \frac{1}{6}|2\rangle + \frac{1}{6}|3\rangle
}{\frac{5}{12}} = \frac{1}{5}|1\rangle + \frac{2}{5}|2\rangle + \frac{2}{5}|3\rangle 
\] \vspace{5px}


\subsection{Operations on Probabilistic States}

We will consider \textit{operations} on multiple systems in probabilistic
states. Returning to the typical setup where we have two systems $X$ and $Y$,
let us consider classical operations on the compound system $(X,Y)$. Any
operation is represented by a stochastic matrix whose rows and columns are
indexed by the Cartesian Product $\Sigma \times \Gamma$. \\

For example, suppose $X$ and $Y$ are bits, and consider the operation: \\

\begin{mainbox}{}
  If $X = 1$, then perform a NOT operation on $Y$. \\ Otherwise do nothing.
\end{mainbox}

This is a deterministic operation known as a \textit{controlled-NOT} or CNOT
operation, where $X$ is the \textit{control} bit that determines whether or not
a NOT operation gets applied to the \textit{target} bit $Y$. Here is the matrix
representation of this operation: 

\[
\begin{pmatrix}
  1 & 0 & 0 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 1 \\
  0 & 0 & 1 & 0 
\end{pmatrix}
\] \vspace{5px}

Its action on the standard basis states: 

\begin{align*}
  &|00\rangle \mapsto |00\rangle \\
  &|01\rangle \mapsto |01\rangle \\
  &|10\rangle \mapsto |11\rangle \\
  &|11\rangle \mapsto |10\rangle 
\end{align*}

Another example is the operation having this description

\begin{subbox}{}
  Perform one of the  following two operations, each with probability
  $\frac{1}{2}$ : 
  \begin{itemize}
    \item[-] Set $Y$ to be equal to $X$ 
    \item[-] Set $X$ to be equal to $Y$
  \end{itemize}
\end{subbox}

The matrix representation is as follows: 

\[
\begin{pmatrix}
  1 & \frac{1}{2} & \frac{1}{2} & 0 \\ 
  0 & 0 & 0 & 0 \\
  0 & 0 & 0 & 0 \\
  0 & \frac{1}{2} & \frac{1}{2} & 1 \end{pmatrix} = \frac{1}{2} \begin{pmatrix}
    1 & 1 & 0 & 0 \\
    0 & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 \\
    0 & 0 & 1 & 1 
  \end{pmatrix} + \frac{1}{2} \begin{pmatrix}
    1 & 0 & 1 & 0 \\
    0 & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 \\
    0 & 1 & 0 & 1
  \end{pmatrix} 
\] \vspace{5px}
The action of this operation on the standard basis vectors: 

\begin{align*}
  &|00\rangle \mapsto |00\rangle \\
  &|01\rangle \mapsto \frac{1}{2}|00\rangle + \frac{1}{2}|11\rangle \\
  &|10\rangle \mapsto \frac{1}{2}|00\rangle  + \frac{1}{2}|11\rangle \\
  &|11\rangle \mapsto |11\rangle 
\end{align*}








\end{document}

